Predicting Water Treatment Plant Chemical Doses

### 2025-12-11 Presentation DSCI445.Rmd
Final edits.

- Created 6 models with different predictors and combinations of interaction terms.

- Used lag variables for temp, ph, conductivity, turbidity

- Dropped Turbidity as it was not a good predictor for FeCL dosing.
  - Conductivity was also dropped eventually due to being similar to Turbidity

- Difficult to determine what factors of the dataset are driving the predictions.

- RMSE and MAE seem to be all over the place dependent on which predictors are selected and the interactions.

@@

__ performed the best out of models tested.

**Insert bar graph/brief conclusions**

@@

**Insert bar graph**

# Conclusion and Possible Next Steps

>>>>

- Created 6 models with different predictors and combinations of interaction terms

- Used lag variables for temp, pH, conductivity, turbidity

- Dropped Turbidity as it was not a good predictor for FeCL dosing

- Difficult to determine what factors of the dataset are driving the predictions

- RMSE and MAE vary widely depending which predictors are selected and the interactions

@@

```{r, fig.height=3.8, fig.width=7}

# - The model with pH & temperature interactions with conductivity performed the best out of models tested.
# - Tree-based models performed the worst, likely due to overfitting.
# - All models diverged from the actual dose by an average of at least 1.4 mg/L.

dosepred_df <- data.frame(Model = c("Random Forest", "Decision Tree", "Cond-pH + Cond-Temp", "Conductivity-Temp", "Temp and pH"),
                          RMSE = c(2.811, 2.756, 1.672, 1.891, 2.350),
                          MAE = c(2.283, 2.484, 1.421, 1.516, 1.941))

# convert to long format
dosepred_long <- dosepred_df |>
  arrange(RMSE) |>
  mutate(Model = factor(Model, levels = unique(Model))) |>
  pivot_longer(
    cols = c(RMSE, MAE),
    names_to = "Metric",
    values_to = "Value"
  )

# Plot generation
ggplot(dosepred_long, aes(x = Model, y = Value, fill = Metric)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.7) +
  geom_text(
    aes(label = sprintf("%.3f", Value)), 
    position = position_dodge(width = 0.8),
    vjust = -0.5, 
    size = 2
  ) +
  labs(
    title = "Dose Prediction Model Comparison (RMSE and MAE)",
    x = "Model",
    y = "Error Value"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```

@@

dosechangepred_df <- data.frame(
  Model = c("LDA", "QDA", "Multinomial", "Random Forest"),
  Accuracy = c(lda_acc$.estimate, qda_acc$.estimate, 
               multi_acc$.estimate, rf_acc$.estimate)
) |> arrange(Accuracy)

ggplot(dosechangepred_df, aes(x = reorder(Model,-Accuracy), y = Accuracy)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = scales::percent(Accuracy, accuracy = 0.1)),
            vjust = -0.3, size = 4) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                     limits = c(0, max(dosechangepred_df$Accuracy) * 1.1)) +
  labs(x = "Model", y = "Accuracy") +
  theme_minimal()

# Conclusion and Next Steps
